---
title: "Differential Gene Expression Analysis with DESeq2"
author: "Mario Farrapo"
date: "2026-03-03"
output: html_document
---

We used RNA-seq data from cell lines treated and untreated with Neutrophil Extracellular Traps (NETs) and performed a differential gene expression analysis with DESeq2 to identify important gene expression differences related to treatment. 

The data were obtained from the GEO repository and are publicly available under code GSE218119.

```{r message=FALSE}

library(tidyverse)
library(DESeq2)
library(ggrepel)

```

### Data import

```{r}

data <- read_table("GSE218119_featurecounts.txt.gz")
data <- data |> column_to_rownames("Geneid") |> select(c(6:11))
counts <- data[, c("VEH1", "VEH2", "VEH3", "NET1", "NET2", "NET3")]
metadata <- data.frame(row.names = colnames(counts), groups = c("VEH", "VEH", "VEH", "NET", "NET", "NET"))

```

### Eliminate genes with an average count of less than 10 in at least 3 samples

```{r}

counts_filtered <- rowSums(counts >= 10) >= 3
counts_filtered <- counts[counts_filtered, ]

```

### Differential expression analysis with DESeq2

```{r message=FALSE}

metadata$groups <- as.factor(metadata$groups)
dds <- DESeqDataSetFromMatrix(countData = counts_filtered,
                              colData = metadata,
                              design = ~ groups)
dds$groups <- relevel(dds$groups, ref = "VEH")
dds <- DESeq(dds)
res <- as.data.frame(results(dds))

```

### Prepare the data for the graph

```{r}

res$diffexpressed <- ifelse(res$log2FoldChange > 1 & res$padj < 0.0001, 
                            "Upregulated",
                            ifelse(res$log2FoldChange < -1 & res$padj < 0.0001,
                                   "Downregulated", "Not significant"))

```

### Volcano plot

```{r warning=FALSE}

(volcano <- ggplot(res, aes(x = log2FoldChange, y = -log10(padj), color = diffexpressed)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Not significant" = "gray")) +
  theme_bw() +
  labs( x = bquote(Log[2]~Fold~Change),
        y = bquote(-Log[10]~padj),
        color = "Status") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.0001), linetype = "dashed", color = "black") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "top"))

```

### Save DEGs table

```{r}

res_table <- filter(res, diffexpressed != "Not significant")
res_table <- rownames_to_column(res_table, "Gene_name")
res_table[,c(2:7)] <- round(res_table[,c(2:7)], 3)
write.table(res_table, "DEGs_table.csv", sep = ";")

```

### Pathway enrichment analysis

```{r message=FALSE}

library(msigdbr)
library(clusterProfiler)
library(enrichplot)

```
```{r message=FALSE, warning=FALSE}

de_vector <- res$stat
names(de_vector) <- rownames(res)
de_vector <- sort(de_vector, decreasing = TRUE)

Hs_hallmark_sets <- msigdbr(species = "Mus musculus", category = "H")

set.seed(123)

gsea_results <- GSEA(
  geneList = de_vector,
  minGSSize = 25,
  maxGSSize = 500,
  pvalueCutoff = 0.05,
  eps = 0,
  seed = TRUE,
  pAdjustMethod = "BH",
  TERM2GENE = dplyr::select(
    Hs_hallmark_sets,
    gs_name,
    gene_symbol
  )
)

gsea_result_df <- data.frame(gsea_results@result)

```

### Plot the graph

```{r warning=FALSE, error=FALSE}

top_enrich <- gsea_result_df %>%
  as_tibble() %>%
  filter(qvalue < 0.05 & NES > 1 | qvalue < 0.05 & NES < -1) %>%
  arrange(desc(NES))

top_enrich$ID <- substr(top_enrich$ID, 10, nchar(top_enrich$ID))

(enrich_plot <- ggplot(top_enrich, aes(x = NES, y = reorder(ID, NES))) +
  geom_segment(aes(x = 0, xend = NES, y = reorder(ID, NES), yend = reorder(ID, NES)),
  color = "black",
  linewidth = 0.7) +
  geom_point(aes(size = rank, color = -log10(qvalue))) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(y="", x=NULL)+
  theme_bw() +
  labs(x = "Normalized Enrichment Score"))

```

### Visualize the enrichment curve of a specific pathway

```{r message=FALSE, warning=FALSE, error=FALSE}

gseaplot2(gsea_results, geneSetID = "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION")

```
